---
title: "Mapping Oil Spills"
description: |
  A short example of working with spatial data.
author:
  - name: iris foxfoot
    url: {}
date: 03-02-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: show code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(gstat)
library(stars)
library(sp)
```

In this post I'm going to do some minor wrangling of spatial data. I'm using oil spills in California as a somewhat depressing example. First we'll read in the data and make an interactive map, where each point represents an oil spill!  

```{r}
#read in the oil spill data, pipe straight into janitor::clean_names(), 
#so that all column names are coder friendly
oil_spill <- read_sf(here("data", "Oil_Spill_Incident_Tracking_%5Bds394%5D-shp")) %>% 
  clean_names()

#read in county shapefile, again cleaning the names. select the names of the 
#counties, the spatial data will come along with it!
ca_counties <- read_sf(here("data", "ca_counties"), layer = "CA_Counties_TIGER2016") %>% 
  clean_names() %>% 
  dplyr::select(name)

#note, check the projection using st_crs(). In this case the oil spill data 
#and the counties data are in the same projection.

#set tmap to interactive mode
tmap_mode("view")

#make an interactive map!
tm_shape(oil_spill) +
  tm_dots("dfgcontrol", legend.show = FALSE)
```

Now we'll make a chloropleth. What's that you say? A chloropleth map is just a map that uses different colors to show an aggregate statistical variable in a given region. In this case we're going to color the counties of California based on how may oil spills were recorded there in 2008. 

```{r}
#join oil spill and counties data for chloropleth
ca_oil_spill <- ca_counties %>% 
  st_join(oil_spill)

#find counts by county
oil_spill_counts <- ca_oil_spill %>% 
  count(name)

#plot chloropleth
ggplot(data = oil_spill_counts) + 
  geom_sf(aes(fill = n), color = "black", size = 0.2) + 
  scale_fill_gradientn(colors = c("white", "orchid","purple")) +
  theme_void() + #theme_void() is great for maps!
  labs(fill = "number of oil spills", 
       title = "California Oil Spills in 2008")
```

It looks like LA County and San Diego County have the most oil spills!

**Data Citation**: California Department of Fish and Game, Office of Spill Prevention and Response. 2008. [https://map.dfg.ca.gov/metadata/ds0394.html](https://map.dfg.ca.gov/metadata/ds0394.html)
