---
title: "Palmetto Binary Logistic Regression"
description: |
  In this post I will demonstrate a simple example of binary logistic regression using data on two palmetto species in Florida.
author:
  - name: iris foxfoot
    url: {}
date: 02-24-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: show code
---
### Intro
In this blog I will use binary logistic regression to predict which species of palmetto observations belong to, then I will compare my results against the actual data, to see how accurate my model is.

Palmettos are small, tropical palms with fan like leaves. In my model, I will use the number of green leaves, height, width, and length to distinguish between two species in Florida: *Serenoa repens*, and *Sabal etonia*.

I'm using data collected by Abrahamson, W.G. 2019, freely available via the Environmental Data Initiative. See the data citation for more information!

click "show code" to see what packages I used in the set up chunk

```{r setup, include = T}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(here)
library(tidyverse)
library(broom)
library(jtools)
library(janitor)
library(kableExtra)
library(knitr)
```

### Data Exploration & Visualization

It's always a good idea to look at your data in a few different ways before you even try any statistical modeling!

```{r, fig.width=8}
#Read in the Data
palmetto <- read_csv(here("palmetto.csv"))

#rename species category because it's annoying
palmetto$species[palmetto$species == "1"] <- "Serenoa repens"
palmetto$species[palmetto$species == "2"] <- "Sabal etonia"

#convert species to factor for analyses
palmetto$species <-as.factor(palmetto$species)

#view it, then make some exploratory graphs
#1 = Serenoa repens, 2 = Sabal etonia

#A ggplot graph showing width and length in the two species
ggplot(data = palmetto, aes(x = width, y = length, fill = species)) + 
  geom_violin() + 
  facet_wrap(~species) + 
  scale_fill_manual(values=c('orange','black')) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0), legend.position = "none") +
  labs(x = "width (cm)", y = "length (cm)",
       caption = "Figure 1: Width and length of Sabal etonia vs Serenoa repens palmettos in South Central Florida (Data: Abrahamson)\nLength is defined as the widest area of the canopy and width is the widest area of the canpopy perpendicular to the \nlength. From this graph you can see that Sabal etonia has a slightly larger mean length.")

#another ggplot exploring height and number of green leaves
ggplot(data = palmetto, aes(x = green_lvs, y = height, color = species)) + 
  geom_point() + 
  scale_color_manual(values=c('orange','black')) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0)) +
  labs(x = "Number of Green Leaves", y = "Height (cm)",
       caption = "Figure 2. Differences in number of green leaves vs height in Sabal Etonia and Serenoa repens (Data: Abrahamson)\nSerenoa repens tends to have a higher number of green leaves. It is difficult to tell if there is a meaningful difference in \nheight")

```

### Binary Logistic Regression

After I've explored the data, I'll use binomial regression to 'predict' what species of palmetto an observation is based on height, width, letght, and the number of green leaves. 

Visualization isn't just for the exploratory phase, it's a great tool for organizing model outputs. Here I've made a nice table to help organize the outputs

```{r}
#levels(palmetto$species)
#seet is first level

#binomial regression
palmetto_blr <- glm(species ~ height + width + length + green_lvs, data = palmetto, family = "binomial")

#call the model to view or summary of the model using summary()

#tidy table
broom::tidy(palmetto_blr) %>%
  mutate(p.value = format(p.value, scientific = TRUE)) %>% 
  kbl(caption = "Palmetto Binary Logistic Regression Model Outputs - Reference Level = Sabal etonia (Data: Abrahamson)") %>% 
  kable_classic()

```

### Model Success

Even with my amazing table, it can be hard to understand the results in this format. Here I've converted the output into a prediction based on a 50% probability threshold, and then again summarized my results in a table. 

You can see that even though exploratory visualization showed a lot of overlap between species variables, the model did a pretty good job!


```{r}
#store in as probabilities
blr_fitted <- palmetto_blr %>% 
  broom::augment(type.predict = "response")

#convert .fitted value to prediction based on 50% probability threshold
sp_prediction_df <- blr_fitted %>% 
  mutate(sp_prediction = 
           if_else(.fitted < .5, "Sabal etonia", "Serenoa repens")) %>% 
  mutate(prediction_correct = 
           if_else(species == sp_prediction, "yes", "no"))

#make an initial table of counts
palmetto_prediction_counts <- sp_prediction_df %>% 
  tabyl(species, prediction_correct)

#table with proportions using adorn()
palmetto_table_proportions <- palmetto_prediction_counts %>% 
  janitor::adorn_percentages() %>% 
  janitor::adorn_pct_formatting(digits = 2) %>% 
janitor::adorn_ns()

#finalized table using kable extra package
palmetto_table_proportions %>%  
  kbl(caption = "Palmetto species and model prediction accuracy (Data: Abrahamson)",
      col.names = c("Species", 
                    "Not Accurately Predicted", 
                    "Accurately Predicted")
      ) %>%
    kable_classic(font ="cambria")
```

### Data Citation
Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5
